### 2015-04-16 implementation-concretization.tex
updates through specs.

\section{The Spack Package Manager}
\label{sec:implementation}
Based on our experiences at LLNL, we have developed
{\it Spack}, the Supercomputing PACKage manager.
Spack is written in Python.  We chose Python because it is
very widely used int he HPC community, and because it provides
flexible scripting capabilities. 
%
Like some prior systems, Spack supports arbitrarily many software installations.
It borrows ideas from Nix; packages are installed in their own prefixes,
and configurations are managed using hashes.
% 
Spack adds several unique capabilities that are essential for HPC:
\begin{enumerate}
\item {\bf Parametric builds} address the matrix problem:
      Packages are parameterized by name, version, architecture, compiler, 
      options, and dependencies.
\item To manage the configuration space, spack provides a novel, 
      recursive {\bf spec syntax} for dependency graphs.
\item Specs support versioned, ABI-incompatible interfaces like MPI through
      {\bf versioned virtual dependencies}.
\item Spack builds with {\bf compiler wrappers} that reduce the burden of build
      consistency on the package author.
\end{enumerate}

\input{implementation-packages}

\subsection{Spack Specs}\label{sec:specs}

  Specs are partial descriptions of full
dependency graphs, but they are concise.  This allows a user to use
only a few parameters to quickly identify a configuration to install,
or to query existing installations and their dependencies.

\todo{1 page}

\subsection{Versioned Virtual Dependencies}
	\todo{.5 page}

\subsection{Abstract \& Concrete Specs}
	\todo{.75 page}

>>>>

\subsection{Abstract Specs \& Concretization}
	\todo{.75 page}

### 2015-04-16 implementation-environment.tex
updates through specs.

\subsection{Packages}

%\subsubsection{Package Files}
\begin{figure}
	\includegraphics[width=\columnwidth]{code/mpileaks.pdf}
	\caption{
		Spack package for the {\tt mpileaks} tool.
		\label{fig:mpileaks}
	}
\end{figure}

Spack comprises many packages, and each describes how to build a software
artifact.  Each package is a class written in pure Python, but Spack uses
a simple, embedded domain-specific language (DSL) to make packaging easier.
Spack's DSL adds directives such as {\tt depends\_on}, {\tt version},
{\tt provides}, and {\tt extends}, in order to add metadata to package classes.

Figure~\ref{fig:mpileaks} shows the package for the \mpileaks tool.
Inside the {\tt MpiLeaks} class, the package provides a text description
and a homepage, as well as 
a download URL.  Next, two {\tt version} directives identify known versions
of the package, and checksums so that they can be downloaded safely.
Below this, three {\tt depends\_on}
directives indicate prerequisites that must be installed before \mpileaks.
Last, each package defines an {\tt install()} method, which contains the
commands used to build the package.

Most package managers fix a single version in a package file, but Spack packages
are {\it parameterized} by version.  In the figure, we see that the {\tt mpileaks}
package specifies two versions.  A user can safely install either by 
specifying it on the command line.  In our experience, however, users
frequently want bleeding-edge versions, and package managers frequently lag 
behind the latest releases.  Spack therefore has a capability to extrapolate
the latest URL, using the {\tt url} attribute as a model\footnote{This works
for consistently named URLs, and it can be specified per-version for packages
with inconsistent naming.}.
The user can ask for an unknown version on the command line and still
successfully download and install it.  Spack can also use the same model to
scrape webpages and find new versions.

\subsubsection{Installation Environment}

%{\bf Reproducibility.}
%\paragraph{Reproducibility}
Spack is intended to build a consistent HPC stack for our multi-user
environment, and reproducible builds are one of our design goals.
Experience at LLNL has shown that it is vexingly difficult to reproduce
a build manually.
%
Many packages used at LLNL have a profusion of build options, and specifying them 
correctly often requires tedious experimentation.  This is due to lack of
build standards and to the diversity of HPC environments.  
For example, in some packages that depend on the {\tt Silo} library,
the {\tt ---with-silo} parameter takes a path to {\tt Silo}'s prefix.
In others, it takes the {\tt include} and {\tt lib} subdirectories,
separated by a comma.
The {\tt install()} method in Spack's package files allows us to record
precise build incantations for later reuse.

@@

>>>>

\subsection{Installation Environment}

%{\bf Reproducibility.}
%\paragraph{Reproducibility}
Spack is intended to build a consistent HPC stack for our multi-user
environment, and reproducible builds are one of our design goals.
Experience at LLNL has shown that it is vexingly difficult to reproduce
a build manually.
%
Many packages used at LLNL have a profusion of build options, and specifying them 
correctly often requires tedious experimentation.  This is due to lack of
build standards and to the diversity of HPC environments.  
For example, in some packages that depend on the {\tt Silo} library,
the {\tt ---with-silo} parameter takes a path to f{\tt Silo}'s prefix.
In others, it takes the {\tt include} and {\tt lib} subdirectories,
separated by a comma.
The {\tt install()} method in Spack's package files allows us to record
precise build incantations for later reuse.

@@

TACC expects the user to run with the same
environment that she built with~\cite{mclay:lmod-tutorial}.

But, at LC, the user is
frequently not the builder of the application, and may not know the particular
incantation required to load the its environment. Moreover, users frequently run more 
than one application in the same job, each built by a separate team, and each with
its own environment. Most LC users are used to running programs with
specific paths; only 10\% use environment modules extensively.

\begin{figure}\centering
	\includegraphics[width=.6\linewidth]{specs/rpaths.pdf}
	\newline
	\verb|spack install mpileaks ^mpich|\newline
	\verb|spack install mpileaks ^openmpi|
	\caption{
		Installing {\tt mpileaks} with MPICH (solid), then with OpenMPI (dashed).  The {\tt dyninst} subtree is
		reused, but {\tt mpileaks} and {\tt callpath} must be rebuilt.
	}
\end{figure}

### 2015-04-16 implementation-packages.tex
intro updates.
updates through specs.

Spack comprises many packages, and each describes how to build a software
artifact.  Each package is a class written in pure Python, but Spack uses
a simple, embedded domain-specific language (DSL) to make packaging easier.
Spack's DSL adds directives such as {\tt depends\_on}, {\tt version},
{\tt provides}, and {\tt extends}, in order to add metadata to package classes.

Figure~\ref{fig:mpileaks} shows the package for the \mpileaks tool.
Inside the {\tt MpiLeaks} class, the package provides a text description
and a homepage, as well as 
a download URL.  Next, two {\tt version} directives identify known versions
of the package, and checksums so that they can be downloaded safely.
Below this, three {\tt depends\_on}
directives indicate prerequisites that must be installed before \mpileaks.
Last, each package defines an {\tt install()} method, which contains the
commands used to build the package.

Most package managers fix a single version in a package file, but Spack packages
are {\it parameterized} by version.  In the figure, we see that the {\tt mpileaks}
package specifies two versions.  A user can safely install either by 
specifying it on the command line.  In our experience, however, users
frequently want bleeding-edge versions, and package managers frequently lag 
behind the latest releases.  Spack therefore has a capability to extrapolate
the latest URL, using the {\tt url} attribute as a model\footnote{This works
for consistently named URLs, and it can be specified per-version for packages
with inconsistent naming.}.
The user can ask for an unknown version on the command line and still
successfully download and install it.  Spack can also use the same model to
scrape webpages and find new versions.

\subsubsection{Installation Environment}

%{\bf Reproducibility.}
%\paragraph{Reproducibility}
Spack is intended to build a consistent HPC stack for our multi-user
environment, and reproducible builds are one of our design goals.
Experience at LLNL has shown that it is vexingly difficult to reproduce
a build manually.
%
Many packages used at LLNL have a profusion of build options, and specifying them 
correctly often requires tedious experimentation.  This is due to lack of
build standards and to the diversity of HPC environments.  
For example, in some packages that depend on the {\tt Silo} library,
the {\tt ---with-silo} parameter takes a path to {\tt Silo}'s prefix.
In others, it takes the {\tt include} and {\tt lib} subdirectories,
separated by a comma.
The {\tt install()} method in Spack's package files allows us to record
precise build incantations for later reuse.

%{\bf Environment isolation.}
\paragraph{Environment isolation}
In addition to command line issues, we
frequently encounter errors due to inconsistencies between the environment of
the package installer and the package user.
%
For example, there are two versions of the {\tt libelf} library used by
LLNL performance tools. One is distributed with RedHat Linux, but another
publicly available version has the same API but an incompatible ABI.
Failure to specify the right version at build time has caused many
inexplicable crashes.
%
Spack manages the build environment by running {\tt install} invocation
in a new process.  It helps packages find dependencies 
correctly, by setting
{\tt PATH}, {\tt PKG\_CONFIG\_PATH}, {\tt CMAKE\_PREFIX\_PATH}, and
{\tt LD\_LIBRARY\_PATH} to include the dependencies of the current build.
These variables are commonly used by build systems to locate dependencies,
and setting them helps to ensure that incorrect libraries are not detected.
The isolated build environment also gives package authors 
free reign to set build-specific environment variables without interfering
with other packages.

%{\bf Compiler wrappers and RPATHs.}
\paragraph{Compiler wrappers and RPATHs}
Finding compilers at build time is not the only obstacle to reproducible
behavior.  As mentioned in Section~\ref{sec:motivation}, it is also important
for binaries to be able to find dependency libraries at {\it runtime}.
One of the most frequent user errors at LC is improper library configuration.
Users frequently do not know what libraries a package was built with, and 
it is difficult for them to construct a suitable {\tt LD\_LIBRARY\_PATH} for
a package that was built by someone else.  Because of frequent support calls,
we typically add {\tt RPATHs} to public software installations, so that paths
to dependencies are embedded in binaries and so that users do not have to know
this information to run installed software correctly.

Spack manages {\tt RPATH} settings and other build policies with
{\it compiler wrappers}. 
In each isolated {\tt install} environment, Spack sets the standard 
environment variables
{\tt CC}, {\tt CXX}, {\tt F77}, and {\tt FC} to point to its own compiler
wrapper scripts.  These variables are used by most build systems to select
C, C++, and Fortran compilers, so they are generally picked up 
automatically\footnote{If builds do not respect {\tt CC}, {\tt CXX}, etc.,
wrappers can typically be added as arguments or inserted into Makefiles
by {\tt install}.}.
When run, the wrappers insert include ({\tt -I}), library ({\tt -L}), and 
{\tt RPATH} ({\tt -Wl,-rpath} or similar) flags into the argument list.
These point to the {\tt include} and {\tt lib} directories of dependency
library installations, where needed headers and libraries are typically located.
After the modified argument list is constructed, the wrappers delegate it
to the real compiler to execute.

Spack's compiler wrappers have a number of useful effects.  First, they allow
Spack to transparently parameterize the compiler for most builds. 
Spack users to easily request a build with a new compiler swapped in.
Second, they enforce the use of {\tt RPATHs} in
installed binaries.  This causes applications built by Spack to run correctly
{\it regardless of the environment}.  Third, because compiler wrappers add 
header and library search paths for dependencies, header and library detection
tests run by most build systems succeed automatically, {\it without}
the need to use special arguments for nonstandard locations.  {\tt configure}
commands in Spack's {\tt install} function can have fewer arguments, and can
be written as they would be for system installs.  This reduces complexity
for package maintainers and enforces consistent, reproducible
build policies across packages.  Finally, because Spack has control over the 
wrappers, package authors can programmatically filter the compiler flags
used by software build systems, a useful last resort when porting to
bleeding-edge platforms or new, esoteric compilers.

\subsubsection{Environment Module Integration}
	\todo{.5 page}

\begin{verbatim}
module disadvantages:
	- lmod solves a lot of problems
	- not on all systems
		- users often run from just a path.
		- not guaranteed that LD_LIBRARY_PATH, etc will
		  be set correctly.

- Lmod assumes build & submit environment must match
	- users often have NO IDEA what environment a package
	  was built with
	- multiple apps may be built with different stacks
	- combining them in a workflow is difficult with modules
		(need to know right incantation)

	- TACC implements a rather complicated mapping back to
	    the build env (Lmod CheckExec)

- Spack solves the matrix problem mentioned in the Lmod
    presentation (slide 98)
	- hierarchy is nasty -- every branch represents a
	  greedy choice
	- might NEED a specific version of ONE dependency
	  that is not supported in the tree
	  => SPACK!

- libmesh
	- boost, petsc, trilinos, grvy, etc.
\end{verbatim}

>>>>

In Spack, packages are build scripts that describe how to build software
artifacts.  Each package is a class written in pure Python to describe
{\it specific} build instructions, and it must extend the {\tt Package}
base class, which implements the more general parts of the build process.
Spack implements a simple, embedded domain-specific language (DSL) to make
packaging easier; it adds special directives such as {\tt depends\_on} and
{\tt version} that add metadata to the class.

Figure~\ref{fig:mpileaks} shows the package for \mpileaks, an LLNL-developed
tool developed for finding leaks in MPI programs.
Inside the {\tt MpiLeaks} class, the package provides a text description
and a homepage, as well as 
a download URL.  Next, two {\tt version} directives identify known versions
of the package, and MD5 checksums ensure they can be downloaded safely.
Below this, three {\tt depends\_on}
directives indicate prerequisites that must be installed before \mpileaks.
Last, each package defines an {\tt install()} method, which contains the
commands used to build the package.  Spack's DSL allows shell
commands to be invoked as Python functions. Here, the {\tt install()} 
method invokes the familiar {\tt configure}, {\tt make}, and
{\tt make install} build commands, as a shell script would.

### 2015-04-16 implementation.tex
updates through specs.

\section{The Spack Package Manager}
\label{sec:implementation}
Based on our experiences at LLNL, we have developed
{\it Spack}, the Supercomputing PACKage manager.
Spack is written in Python.  We chose Python because it is
very widely used int he HPC community, and because it provides
flexible scripting capabilities. 
%
Like some prior systems, Spack supports arbitrarily many software installations.
It borrows ideas from Nix; packages are installed in their own prefixes,
and configurations are managed using hashes.
% 
Spack adds several unique capabilities that are essential for HPC:
\begin{enumerate}
\item {\bf Parametric builds} address the matrix problem:
      Packages are parameterized by name, version, architecture, compiler, 
      options, and dependencies.
\item To manage the configuration space, spack provides a novel, 
      recursive {\bf spec syntax} for dependency graphs.
\item Specs support versioned, ABI-incompatible interfaces like MPI through
      {\bf versioned virtual dependencies}.
\item Spack builds with {\bf compiler wrappers} that reduce the burden of build
      consistency on the package author.
\end{enumerate}

\input{implementation-packages}

\subsection{Spack Specs}\label{sec:specs}

  Specs are partial descriptions of full
dependency graphs, but they are concise.  This allows a user to use
only a few parameters to quickly identify a configuration to install,
or to query existing installations and their dependencies.

\todo{1 page}

\subsection{Versioned Virtual Dependencies}
	\todo{.5 page}

\subsection{Abstract \& Concrete Specs}
	\todo{.75 page}

>>>>

\section{The Spack Package Manager}
\label{sec:implementation}
Based on our experiences at LLNL, we have developed
{\it Spack}, the Supercomputing PACKage manager.
Spack is written in Python.  We chose Python because it is
very widely used int he HPC community, and because it provides
flexible scripting capabilities. 
%
Like some prior systems, Spack supports arbitrarily many software installations.
It borrows ideas from Nix; packages are installed in their own prefixes,
and configurations are managed using hashes.

\noindent
Spack adds several essential capabilities for HPC:
\begin{enumerate}
\item {\bf Parametric builds} address the combinatorial problem:
      Packages are parameterized by name, version, architecture, compiler, 
      options, and dependencies.
\item To manage the configuration space, spack provides a novel, 
      recursive {\bf spec syntax} for dependency graphs.
\item Specs support versioned, ABI-incompatible interfaces like MPI through
      {\bf versioned virtual dependencies}.
\item Spack builds with {\bf compiler wrappers} that reduce the burden of build
      consistency on the package author.
\end{enumerate}

\input{implementation-packages}
\input{implementation-specs}
\input{implementation-virtual-deps}
\input{implementation-concretization}

\input{implementation-environment}

### 2015-04-16 motivation.tex
updates through specs.

This section describes existing tools for managing HPC 
software and motivates the need for a system like Spack.

\paragraph{Meta-build Systems}

{\it Meta-build systems} such as Contractor, WAF, and
MixDown~\cite{amundson:contractor,epperly+:mixdown,nagy:waf} are closely
related to package managers, but their focus is on ensuring that a single
project builds correctly along with all of its dependencies.  They do not address
larger, multi-user system administration issues such as versioning, naming
conventions, rapid multi-configuration installation, or maintaining many
divergent software stacks.

\paragraph{Traditional package managers}
Package managers automate the installation of complex sets of software packages.
{\it Binary package managers} such as RPM, yum, and  
APT~\cite{foster+:rpm03,silva:apt01,yum} are integrated with most
operating system (OS) distributions, and they are used to ensure that dependencies
are installed before packages that require them.
These systems largely solve the problem of managing a single software
stack, which works well for the baseline OS and drivers, which are 
common to all applications on a system.
However, they make the assumption that there is only a single version
of each package, and most install packages in a single, inflexible location.
To install multiple configurations, the user must create a custom, combinatoric
naming scheme to avoid conflicts. They typically require root 
privileges and do not optimize for specific hardware.

{\it Port systems} such as Gentoo, BSD Ports, MacPorts, and 
Homebrew~\cite{bsdports,groffen:gentoo-prefix,homebrew,macports,thiruvathukal:gentoo04}
allow packages to be built on demand, but most suffer from 
the same versioning and naming issues as traditional package managers.
Some allow multiple versions to be installed in the same
prefix~\cite{groffen:gentoo-prefix}, but again the burden is on package
creators to manage conflicts. This restricts installations to only a few
configurations.

@@

Looking again at the conventions in Table~\ref{tab:naming-conventions},
we see that most sites use some combination of architecture, compiler version,
package name, package version, and a custom (up to the author, sometimes
encoded) build identifier.  TACC and many other sites also include the MPI
version in the path. None of these naming conventions is defined for
the entire configuration space, and none defines a way to create new versions
from arbitrary combinations of dependencies.  MPI is explicitly called out,
but it is only one of many dependencies that go into a build.
Sites have made a conscious choice to divide the directory hierarchy
according to the most common software packages, but they ignore the rest.
Frequently, desired configurations do not fit the local convention, and
staff must install them in nonstandard locations.

@@

Diverse software versions not only present problems for build and installation; 
they also complicate the runtime environment. When run, an executable
must determine the location of its dependency libraries, or it will fail to run.
This is typically not a problem for statically linked executables, but modern
operating systems make extensive use of dynamic linking, which has many 
benefits for software tools and maintenance.
By default, the dynamic loader on most systems is configured to search only
system library paths such as {\tt /lib}, {\tt /usr/lib}, and
{\tt /usr/local/lib}.  If binaries are installed to other locations, the
{\it user} who runs the program must typically add dependency library paths to
{\tt LD\_LIBRARY\_PATH} (or a similar environment variable on other OSes).
Most HPC sites address this problem using {\it environment modules}, which allow
users to dynamically ``load'' and ``unload'' settings for search paths like
{\tt LD\_LIBRARY\_PATH}, {\tt PATH}, {\tt MANPATH}, etc.

Environment modules date back to 1991, and there are many implementations 
available~\cite{dotkit,furlani+:lisa91,furlani+:lisa96,mclay:lmod,mclay:lmod-tutorial}.
The most advanced implementations, such as Lmod~\cite{mclay:lmod,mclay:lmod-tutorial},
provide software hierarchies similar to the naming conventions in 
Table\ref{tab:naming-conventions}, and they allow users to quickly load a software stack
if they know which one is required.  TACC expects the user to run with the same
environment that she built with~\cite{mclay:lmod-tutorial}.  But, at LC, the user is
frequently not the builder of the application, and may not know the particular
incantation required to load the its environment. Moreover, users frequently run more 
than one application in the same job, each built by a separate team, and each with
its own environment. Most LC users are used to running programs with
specific paths; only 10\% use environment modules extensively.

@@

Existing tools do not enforce {\tt RPATHs}, leaving them to package authors.
This may lead to erroneous runs.

>>>>

\paragraph{Meta-build Systems}
{\it Meta-build systems} such as Contractor, WAF, and
MixDown~\cite{amundson:contractor,epperly+:mixdown,nagy:waf} are closely
related to package managers, but their focus is on ensuring that a single
{\it project} builds correctly along with all of its dependencies.  They do not address
larger, multi-user system administration issues such as versioning, naming
conventions, rapid multi-configuration installation, or maintaining many
divergent software stacks.

\paragraph{Traditional package managers}
Package managers automate the installation of complex sets of software packages.
{\it Binary package managers} such as RPM, yum, and  
APT~\cite{foster+:rpm03,silva:apt01,yum} are integrated with most
operating system (OS) distributions, and they are used to ensure that dependencies
are installed before packages that require them.
These tools largely solve the problem of managing a {\it single} software
stack.  This works well for the baseline OS and drivers, which are 
common to all applications on a system.
These tools make the assumption that there is only a single version
of each package, and most install packages in a single, inflexible location.
To install multiple configurations, the user must create a custom, combinatoric
naming scheme to avoid conflicts. They typically require root 
privileges and do not optimize for specific hardware.

{\it Port systems} such as Gentoo, BSD Ports, MacPorts, and 
Homebrew~\cite{bsdports,groffen:gentoo-prefix,homebrew,macports,thiruvathukal:gentoo04}
build packages from source instead of installing from a pre-built binary.
Most existing port systems suffer from 
the same versioning and naming issues as traditional package managers.
Some allow multiple versions to be installed in the same
prefix~\cite{groffen:gentoo-prefix}, but again the burden is on package
creators to manage conflicts. This restricts installations to only a few
configurations.

@@

From the conventions in Table~\ref{tab:naming-conventions},
we see that most sites use some combination of architecture, compiler version,
package name, package version, and a custom (up to the author, sometimes
encoded) build identifier.  TACC and many other sites also explicitly
include the MPI version in the path. MPI is explicitly called out
because it is one of the most common software packages for HPC.
However, it is only one of many dependencies that go into a build.
None of these naming conventions covers the entire configuration
space, and none has a way to represent, e.g., two builds that are identical
save for the version of a particular dependency library.  In our experience
at LLNL, naming conventions like these have not succeeded because there
are more desired configurations than we can represent with a practical
directory hierarchy. Staff frequently install nonconforming packages 
in nonstandard locations with ambiguous names.

@@

Diverse software versions not only present problems for build and installation; 
they also complicate the runtime environment. When launched, an executable
must determine the location of its dependency libraries, or it will not run.
Statically linked binaries do not have this issue, but modern
operating systems make extensive use of dynamic linking.
By default, the dynamic loader on most systems is configured to search only
system library paths such as {\tt /lib}, {\tt /usr/lib}, and
{\tt /usr/local/lib}.  If binaries are installed to other locations, the
{\it user} who runs the program must typically add dependency library paths to
{\tt LD\_LIBRARY\_PATH} (or a similar environment variable) so that the loader
can find them.

Many HPC sites address this problem using {\it environment modules}, which allow
users to dynamically ``load'' and ``unload'' such settings using simple commands.
Environment emerged in 1991, and there are many implementations 
available~\cite{dotkit,furlani+:lisa91,furlani+:lisa96,mclay:lmod,mclay:lmod-tutorial}.
The most advanced implementations, such as Lmod~\cite{mclay:lmod,mclay:lmod-tutorial},
provide software hierarchies similar to the naming conventions in 
Table~\ref{tab:naming-conventions}. They allow users to quickly load a software stack
if they know which one is required.

@@

%Existing tools do not enforce {\tt RPATHs}, leaving them to package authors.
%This may lead to erroneous runs.
